What is happening on Twitter? A framework for student research projects with tweets

### 2020-06-05 Rmd/jse-special-issue-2020.Rmd
updated per comments of June 4 ms

In our experience, student interests vary, and students' ability to articulate research goals 
may be limited. An initial brainstorming session may clarify their interests and encourage 
them to think critically about goals under the time constraints of their academic schedules. 
Additionally, we anticipate that sharing completed student project reports will guide student 
thinking about the scope of possible projects.

We mentored one student whose interest in financial time series and tweet sentiment analysis 
guided her project. A second student formulated a project around event detection from tweet 
time series.

@@

From our collection of tweets, we chose to analyze those sent on three consecutive 
days from May 24, 2020 to May 26, 2020. We wanted to see if we could use text analysis 
tools to detect a transient change in topics for Memorial Day (May 25, 2020). To minimize the computing requirements, we 
limited our analysis to tweets sent during a five-minute period (12:00pm to 12:05pm Eastern time) every day. However, our methods are appropriate for much larger data sets.
We then submitted API queries to Twitter
to get the full content of tweets, including the tweet text.
In supplementary files, we provide the R code that we used to 
query the Twitter API to obtain full tweet content.

Twitter API use agreements forbid users from sharing complete API query results. 
However, Twitter permits users to share tweet identification numbers. A user can 
then query a Twitter API to obtain complete tweet data. In our experience, this 
process is incomplete; that is, many tweets submitted to the Twitter API return 
no data. Additionally, on repeated querying of the API, different sets of tweets 
return data. This complicates our goal of making all analyses computationally 
reproducible and motivates our decision to share the tweet IDs of those tweets that 
we actually analyzed.

@@

We used functions from the `rtweet` R package to parse tweet JSON into a data frame [@rtweet-package]. 
We then divided tweet text into words with functions from the `tidytext` R package [@tidytext]. We 
discarded commonly used "stop words" and emojis.

@@

Inference for latent Dirichlet allocation model is performed by either 
sampling from the posterior distribution or through variational methods. Researchers
have devised a variety of Gibbs sampling techniques for these models. 
Variational methods, while using approximations to the posterior distribution, offer the 
advantage of computational speed [@blei2017variational]. We used variational methods below.

@@

We sought to validate our hypothesis that we could detect a social 
media event by examining tweet topic content at distinct time periods. 
As a proof of principle of our event detection strategy, we chose to 
analyze tweets before, during, and after Memorial Day (May 25, 2020). We fitted latent Dirichlet allocation models for each of three
distinct five-minute periods. The first period began at noon Eastern time on May 24, 2020. Subsequent time periods started 
24 and 48 hours later. 
We defined each time period to be a single collection, or corpus, of tweets.

@@

We examined student learning through both formative and summative assessments.
We conducted formative assessments through weekly discussions with students.
In these discussions, we developed action items to advance research progress and overcome challenges.

We summatively assessed student achievement at the end of the academic year. Both
students wrote a thesis and presented a poster to our statistics department. 
We asked questions at the poster session to probe student understanding and critically
evaluated the theses.

@@

In future iterations, we will use a written rubric to grade student theses (Table \@ref(tab:summative)). We'll
share the rubric with our students at the start of the academic year.

@@

Tweet collections over time are a rich, large, authentic data set that offer 
many opportunities for student research projects. We provided 
instructions to enable readers to 
establish their own tweet collections. We also presented details for one 
mentored research project that made use of our stored tweets.

Our mentoring in data science aligns with others' calls to reconsider training in statistics
and data science [@nolan2010computing;@carver2016guidelines]. @hicks2018guide argue for
incorporating three concepts into data science classes: computing, connecting and creating. 
They use the terms "connecting" and "creating" to describe the processes of applying 
quantitative methods to real data and real research questions and of research question 
formulation, respectively. Our tweet analysis projects offer students opportunities in all
three skills sets. Our students first formulated research questions, then collected and 
analyzed data to address the questions. Throughout the projects, students 
drew heavily on computing, both to acquire data and to analyze it.

>>>>

In our experience, student interests vary, and students' ability to articulate research goals 
may be limited. An initial brainstorming session may clarify their interests and encourage 
them to think critically about goals under the time constraints of their academic schedules. 
Additionally, we anticipate that sharing completed student project reports will guide student 
thinking about the scope of possible projects (Supplemental files).

@@

We analyzed tweets sent on three consecutive 
days from May 24, 2020 to May 26, 2020. We wanted to see if we could detect a transient change in topics for Memorial Day (May 25, 2020). To minimize the computing requirements, we 
limited our analysis to tweets sent during a five-minute period (12:00pm to 12:05pm Eastern time) every day. However, our methods are also appropriate for much larger data sets.

Twitter API use agreements forbid users from sharing complete API query results. 
However, Twitter permits users to share tweet identification numbers. With a tweet 
identification number, a user may query a Twitter API to obtain complete tweet data. In our 
experience, this 
process is incomplete; that is, many tweets submitted to the Twitter API return 
no data. Additionally, on repeated querying of the API, different sets of tweets 
return data. This complicates our goal of making all analyses computationally 
reproducible and motivates our decision to share the tweet IDs of those tweets that 
we actually analyzed (Supplemental files).

@@

The next task is to wrangle the tweet JSON into a data structure suitable for latent Dirichlet allocation modeling. 
We used functions from the `rtweet` R package to parse tweet JSON into a data frame [@rtweet-package]. 
We then divided tweet text into words with functions from the `tidytext` R package [@tidytext]. We 
discarded commonly used "stop words" and emojis.

@@

Inference for latent Dirichlet allocation models is performed by either 
sampling from the posterior distribution or through variational methods. Researchers
have devised a variety of Gibbs sampling techniques for these models [@porteous2008fast]. 
Variational methods, while using approximations to the posterior distribution, offer the 
advantage of computational speed [@blei2017variational]. We used variational methods below.

@@

We sought to validate our hypothesis that we could detect a social 
media event by examining tweet topic content at distinct time periods. 
As a proof of principle of our event detection strategy, we analyzed tweets before, during, and 
after Memorial Day (May 25, 2020). We fitted latent Dirichlet allocation models for each of three
distinct five-minute periods. The first period began at noon Eastern time on May 24, 2020. Subsequent time periods started 
24 and 48 hours later. 
We defined each time period to be a single collection, or corpus, of tweets.

@@

We examined student learning with both formative and summative assessments.
We conducted formative assessments through weekly discussions with students.
In these discussions, we developed action items to advance research progress and overcome challenges. We summatively assessed student achievement at the end of the academic year. Both
students wrote a thesis and presented a poster to our statistics department. 
We asked questions at the poster session to probe student understanding and critically
evaluated the theses.

@@

With future students, we will use a written rubric to evaluate theses (Table \@ref(tab:summative)). We'll
share the rubric with our students at the start of the academic year.

@@

Tweet collections over time are a rich, large, authentic data set that offer 
many opportunities for student research projects. We provided 
instructions to enable readers to 
establish their own tweet collections. We also presented details for one analysis strategy.

Our mentoring in data science aligns with others' calls to reconsider the role of computing in statistics
and data science [@nolan2010computing;@carver2016guidelines]. @hicks2018guide argue for
incorporating three concepts into data science classes: computing, connecting and creating. 
They use the terms "connecting" and "creating" to describe the processes of applying 
quantitative methods to real data and real research questions and of research question 
formulation, respectively. Our tweet analysis projects offer students opportunities in all
three skills sets. Our students first formulated research questions, then collected and 
analyzed data to address the questions. Throughout the projects, students 
drew heavily on computing, both to acquire data and to analyze it.
